---
title: "ggsci 3.2.0: new color palettes from Observable, Bootstrap, and Tailwind CSS"
author: "Nan Xiao"
date: "2024-07-06T00:20:00"
slug: ggsci-3-2-0
categories: []
tags:
  - R
  - R packages
  - ggplot2
  - ggsci
  - color scales
  - data visualization
  - Observable
  - Bootstrap
  - Tailwind CSS
meta_img: "image/europeana--HoDFUBpLrA-unsplash.jpg"
description: >
  ggsci 3.2.0 brings the Observable 10, Bootstrap, and Tailwind CSS color
  palettes to R. Create more effective and visually appealing ggplot2
  visualizations using these modern, accessible color scales.
---

```{r, include=FALSE}
knitr::knit_hooks$set(pngquant = knitr::hook_pngquant)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>",
  dpi = 96,
  fig.retina = 2,
  fig.width = 10.6667,
  fig.height = 3.3334,
  fig.align = "center",
  out.width = "100%",
  pngquant = "--speed=1 --quality=90"
)
```

![Villa in Zurich, perspective view. From *"Moderne Architektur"* (Modern Architecture), edited by Lambert & Stahl, Stuttgart, 1891. Scan from [Europeana](https://unsplash.com/photos/-HoDFUBpLrA).](/image/europeana--HoDFUBpLrA-unsplash.jpg)

I am delighted to announce the release of [ggsci](https://nanx.me/ggsci/) 3.2.0.
The R package ggsci was first released in 2016.
It offers a range of ggplot2 color scales drawn from various sources,
including scientific publications, data visualization tools,
sci-fi movies, and television series.

To install ggsci from CRAN, use:

```r
install.packages("ggsci")
```

As a follow-up to the [ggsci 3.1.0 release](https://nanx.me/blog/post/ggsci-3-1-0/),
ggsci 3.2.0 introduces three new color scales and continues to improve the
package engineering details.

## Observable 10

The Observable 10 palette was
[officially announced](https://observablehq.com/blog/crafting-data-colors)
in April 2024. It was designed as the new default categorical color scheme for
Observable Plot, the JavaScript library for tabular data visualization.
To me, this new color palette offers a fresh and clean aesthetic while
maintaining accessibility and robustness for data visualization purposes.

```{r}
#| fig-ggsci-observable-color-palette,
#| fig.cap="The Observable 10 palette added in ggsci 3.2.0.",
#| echo=FALSE,
#| dpi=300,
#| fig.width=6,
#| fig.height=3.2,
#| out.width="100%"
color_palette <- ggsci::pal_observable()(10)

plot(
  NULL,
  xlim = c(0, 10), ylim = c(0, 3),
  xlab = "", ylab = "",
  xaxt = "n", yaxt = "n",
  bty = "n"
)

for (i in 1:10) {
  rect(i - 1, 2, i, 3, col = color_palette[i], border = NA)
}

text(1:10 - 0.19, rep(1.8, 10), labels = color_palette, srt = 0, adj = 1, xpd = TRUE, cex = 0.3)
```

In ggsci, this palette is made available via the new color scale functions
[`scale_color_observable()` and `scale_fill_observable()`](https://nanx.me/ggsci/reference/scale_observable.html).
Here is a ggplot2 example that uses this color palette to visualize five categories.

```{r}
p1 <- ggplot2::ggplot(
  subset(ggplot2::diamonds, carat >= 2.2),
  ggplot2::aes(x = table, y = price, colour = cut)
) +
  ggplot2::geom_point(alpha = 0.7) +
  ggplot2::geom_smooth(method = "loess", alpha = 0.05, linewidth = 1, span = 1) +
  ggplot2::theme_bw()
p2 <- ggplot2::ggplot(
  subset(ggplot2::diamonds, carat > 2.2 & depth > 55 & depth < 70),
  ggplot2::aes(x = depth, fill = cut)
) +
  ggplot2::geom_histogram(colour = "black", binwidth = 1, position = "dodge") +
  ggplot2::theme_bw()
```

```{r}
#| fig-ggsci-observable-color-scale,
#| fig.cap="ggplot2 example using the Observable 10 color scales.",
p1 <- p1 + ggsci::scale_color_observable()
p2 <- p2 + ggsci::scale_fill_observable()
gridExtra::grid.arrange(p1, p2, ncol = 2)
```

## Bootstrap and Tailwind CSS palettes

It has been quite a while since we expanded continuous color scales in ggsci.
For a new start, I have chosen to implement the default color systems from
Bootstrap 5 and Tailwind CSS 3, the two most popular front-end frameworks.
These palettes have been refined over years of development, making them
possible choices for data visualization.

Using ColorBrewer's terminology, these color systems contain multiple
single-hue sequential color schemes.
They are suitable for representing values in ordered or numeric data,
such as heatmap visualizations.

In ggsci, these are made available via
[`scale_color_bs5()`/`scale_fill_bs5()`](https://nanx.me/ggsci/reference/scale_bs5.html)
and [`scale_color_tw3()`/`scale_fill_tw3()`](https://nanx.me/ggsci/reference/scale_tw3.html).
We can easily visualize them using
[`colorspace::swatchplot()`](https://colorspace.r-forge.r-project.org/reference/swatchplot.html).

```{r}
#| fig-ggsci-bootstrap-color-palette,
#| fig.cap="The 11 Bootstrap 5 continuous palettes added in ggsci 3.2.0.",
#| echo=FALSE,
#| dpi=300,
#| fig.width=6,
#| fig.height=4,
#| out.width="100%"
colorspace::swatchplot(
  ggsci:::ggsci_db$bs5,
  nrow = 11,
  line = 2.5,
  cex = 0.8,
  border = "transparent",
  sborder = "transparent"
)
```

```{r}
#| fig-ggsci-tailwind-color-palette,
#| fig.cap="The 22 Tailwind CSS continuous palettes added in ggsci 3.2.0.",
#| echo=FALSE,
#| dpi=300,
#| fig.width=6,
#| fig.height=6,
#| out.width="100%"
colorspace::swatchplot(
  ggsci:::ggsci_db$tw3,
  nrow = 22,
  line = 2.5,
  cex = 0.8,
  border = "transparent",
  sborder = "transparent"
)
```

## Other improvements

As the package grows, it is even more important to keep things light
so that we have enough space to expand to even more color scales comfortably.

I first
[moved the internal color palette data](https://github.com/nanxstats/ggsci/pull/43)
out of the binary file `R/sysdata.rda` and moved it into `R/palettes.R` instead.
This change makes everything plain text
(which is [unreasonably effective](https://www.youtube.com/watch?v=WgV6M1LyfNY)).
I hope this simplifies third-party contributions by eliminating the
tedious `.rda` file generation step with the previous R script in `data-raw/`.

The other set of optimizations was applied to the continuous color scale
code examples. Besides removing the soft dependency reshape2 from the code,
I fine-tuned the layout with a more compact grid to reduce the output
image size in [#46](https://github.com/nanxstats/ggsci/pull/46), reducing the
package size concerns described in the
[previous post](https://nanx.me/blog/post/rpkgs-pngquant-ragg/).

## What's next?

I'm open to new ideas on the next color palettes to add.
If you have a suggestion, please submit them by
[creating an issue](https://github.com/nanxstats/ggsci/issues).
