---
title: "ggsci 4.0.0: 400+ new color palettes"
author: "Nan Xiao"
date: "2025-09-25T23:36:00"
slug: ggsci-4-0-0
categories: []
tags:
  - R
  - Python
  - ggplot2
  - plotnine
  - ggsci
  - color palettes
  - color scales
  - data visualization
  - iterm2-color-schemes
  - R packages
  - Python packages
meta_img: "image/alisa-anton-JhxGkGgd3Sw-unsplash.jpg"
description: >
  ggsci 4.0.0 is here, with over 400 new color palettes from Primer,
  Atlassian, and iTerm2-Color-Schemes to bring more options to your
  ggplot2 data visualizations.
---

```{r, include=FALSE}
knitr::knit_hooks$set(pngquant = knitr::hook_pngquant)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>",
  dpi = 96,
  fig.retina = 2,
  fig.width = 10.6667,
  fig.height = 3.3334,
  fig.align = "center",
  out.width = "100%",
  pngquant = "--speed=1 --quality=90"
)
```

![Cozy autumn vibes: a warm drink, a good book, and colorful leaves.
Photo by [Alisa Anton](https://unsplash.com/photos/JhxGkGgd3Sw).
](/image/alisa-anton-JhxGkGgd3Sw-unsplash.jpg)

After the [last update](https://nanx.me/blog/post/ggsci-3-2-0/) in June 2024,
I am excited to announce the release of [ggsci](https://nanx.me/ggsci/) 4.0.0
and [ggsci for Python](https://nanx.me/py-ggsci/) 1.0.0.
You can install the R package ggsci from CRAN:

```r
install.packages("ggsci")
```

And install the Python package from PyPI:

```bash
pip install ggsci
```

This update brings over 400 new discrete color palettes for ggplot2 and
plotnine visualizations. With this many new options, I _almost surely_ will not
have to worry about running out of palettes for producing my plots again.

## New color scales

This release introduces three new discrete color scales, with the underlying
palettes sourced from popular design systems and terminal color schemes.

### Primer

Primer is GitHub's design system. This color scale includes the 17 mark colors
from Primer's [data visualization guide
](https://primer.style/product/ui-patterns/data-visualization/).

```{r, echo=TRUE}
#| fig-ggsci-primer-color-scale,
#| fig.cap="ggplot2 examples using the Primer color scale.",
p1 <- ggsci::example_scatterplot() + ggsci::scale_color_primer()
p2 <- ggsci::example_barplot() + ggsci::scale_fill_primer()
gridExtra::grid.arrange(p1, p2, ncol = 2)
```

### Atlassian Design System

The Atlassian Design System color scale is based on their [chart colors for data
visualization](https://atlassian.design/foundations/color-new/data-visualization-color).
It offers a specific sequence of 8 colors for categorical data.

```{r, echo=TRUE}
#| fig-ggsci-atlassian-color-scale,
#| fig.cap="ggplot2 examples using the Atlassian Design System color scale.",
p1 <- ggsci::example_scatterplot() + ggsci::scale_color_atlassian()
p2 <- ggsci::example_barplot() + ggsci::scale_fill_atlassian()
gridExtra::grid.arrange(p1, p2, ncol = 2)
```

### iTerm

The iTerm color scale includes all current **435 color schemes** from the
[iterm2-color-schemes](https://github.com/mbadolato/iTerm2-Color-Schemes) project.
Each palette provides 6 colors with two possible variants: normal and bright.
To help you explore all the options, I created a Quarto microsite where you
can preview every palette: <https://nanx.me/ggsci-iterm/>.
The page is quite long, so be sure to use the independently scrollable table of
contents on the sidebar.

```{r, echo=TRUE}
#| fig-ggsci-iterm-color-scale,
#| fig.cap="ggplot2 examples using the iTerm color scale with the Rosé Pine palette---a favorite Neovim color scheme used by @ThePrimeagen.",
p1 <- ggsci::example_scatterplot() + ggsci::scale_color_iterm("Rose Pine")
p2 <- ggsci::example_barplot() + ggsci::scale_fill_iterm("Rose Pine")
gridExtra::grid.arrange(p1, p2, ncol = 2)
```

So, why add terminal color schemes?

For months, I have been using the [Ghostty](https://ghostty.org/) terminal
emulator, which features built-in themes from the iterm2-color-schemes project.
I thought it would be cool to use these beautiful themes in data visualizations.
While not all are suitable for plotting, many are designed with decent contrast
and vibes. This inspired me to bring them to ggsci, so you can easily apply
them to your ggplot2 and plotnine pipelines.

## Customizing color scales

As a reminder, the R package includes a
[FAQ vignette](https://nanx.me/ggsci/articles/ggsci-faq.html)
that addresses the most common questions from real ggsci users, such as
"how to handle more data categories than available colors" or "how to customize
the color order in a palette". There, you will find practical recipes to
get your work done quickly.
