---
title: "ggsci 3.0.0"
author: "Nan Xiao"
date: "2023-03-08T11:30:00"
slug: ggsci-3-0-0
categories: []
tags:
  - R
  - ggplot2
  - color scales
meta_img: "image/rodion-kutsaiev-0d2zObuClQI-unsplash.jpg"
description: >
  ggsci 3.0.0 brings 7 new color palettes, multiple improvements,
  and a new FAQ vignette.
---

```{r, include=FALSE}
knitr::knit_hooks$set(pngquant = knitr::hook_pngquant)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>",
  dev = "ragg_png",
  dpi = 96,
  fig.retina = 2,
  fig.width = 10.6667,
  fig.height = 3.3334,
  fig.align = "center",
  out.width = "100%",
  pngquant = "--speed=1 --quality=50"
)
```

![3D render from [Rodion Kutsaiev](https://unsplash.com/photos/0d2zObuClQI).](/image/rodion-kutsaiev-0d2zObuClQI-unsplash.jpg)

I am delighted to announce the release of [ggsci](https://nanx.me/ggsci/) 3.0.0.
ggsci provides a collection of color palettes inspired by colors used in
scientific journals, data visualization libraries, science fiction movies,
and TV shows.

You can install it from CRAN with:

```r
install.packages("ggsci")
```

This version brings 7 new color palettes under 3 color scales
(COSMIC, Flat UI, Frontiers), multiple improvements, and a new FAQ vignette,
giving you more options to create effective data visualizations using ggplot2.

## New color palettes

### COSMIC

Three COSMIC color palettes are added in ggsci 3.0.0.
These color palettes are inspired by the colors used in projects from the
[Catalogue Of Somatic Mutations in Cancers (COSMIC)](https://cancer.sanger.ac.uk/cosmic) and are contributed by [Joshua H. Cook](https://github.com/jhrcook).

```{r}
library("ggsci")
library("ggplot2")
library("gridExtra")

data("diamonds")

p1 <- ggplot(
  subset(diamonds, carat >= 2.2),
  aes(x = table, y = price, colour = cut)
) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", alpha = 0.05, linewidth = 1, span = 1) +
  theme_bw()
p2 <- ggplot(
  subset(diamonds, carat > 2.2 & depth > 55 & depth < 70),
  aes(x = depth, fill = cut)
) +
  geom_histogram(colour = "black", binwidth = 1, position = "dodge") +
  theme_bw()
```

The hallmarks (light) palette:

```{r, cosmic-hallmarks-light}
p1_cosmic_hallmarks_light <- p1 + scale_color_cosmic("hallmarks_light")
p2_cosmic_hallmarks_light <- p2 + scale_fill_cosmic("hallmarks_light")
grid.arrange(p1_cosmic_hallmarks_light, p2_cosmic_hallmarks_light, ncol = 2)
```

The hallmarks (dark) palette:

```{r, cosmic-hallmarks-dark}
p1_cosmic_hallmarks_dark <- p1 + scale_color_cosmic("hallmarks_dark")
p2_cosmic_hallmarks_dark <- p2 + scale_fill_cosmic("hallmarks_dark")
grid.arrange(p1_cosmic_hallmarks_dark, p2_cosmic_hallmarks_dark, ncol = 2)
```

The signature (substitutions) palette:

```{r, cosmic-signature}
p1_cosmic_signature <- p1 + scale_color_cosmic("signature_substitutions")
p2_cosmic_signature <- p2 + scale_fill_cosmic("signature_substitutions")
grid.arrange(p1_cosmic_signature, p2_cosmic_signature, ncol = 2)
```

### Flat UI

Three color palettes from [Flat UI Colors 2](https://flatuicolors.com/)
are included in ggsci 3.0.0, contributed by
[Clara Jégousse](https://github.com/clarajegousse).

The default palette:

```{r, flatui-default}
p1_flatui <- p1 + scale_color_flatui()
p2_flatui <- p2 + scale_fill_flatui()
grid.arrange(p1_flatui, p2_flatui, ncol = 2)
```

The "flattastic" palette:

```{r, flatui-flattastic}
p1_flatui <- p1 + scale_color_flatui("flattastic")
p2_flatui <- p2 + scale_fill_flatui("flattastic")
grid.arrange(p1_flatui, p2_flatui, ncol = 2)
```

The "aussie" palette:

```{r, flatui-aussie}
p1_flatui <- p1 + scale_color_flatui("aussie")
p2_flatui <- p2 + scale_fill_flatui("aussie")
grid.arrange(p1_flatui, p2_flatui, ncol = 2)
```

### Frontiers

Clara also contributed one color palette with inspirations from
[Frontiers](https://www.frontiersin.org/):

```{r, frontiers}
p1_frontiers <- p1 + scale_color_frontiers()
p2_frontiers <- p2 + scale_fill_frontiers()
grid.arrange(p1_frontiers, p2_frontiers, ncol = 2)
```

## Other notable improvements

A new [FAQ vignette](https://nanx.me/ggsci/articles/ggsci-faq.html) is added,
with answers to two most frequently asked questions:

- A simple solution to interpolate the color palettes when the data has more
  categories than the number of colors in a discrete color scale.
- A method for applying color scales consistently for multiple ggplot2 plots by
  setting global options.

The output figure size is now reduced in vignettes and `README.Rmd` by
switching to the [ragg](https://ragg.r-lib.org/) PNG device, plus using
[pngquant](https://pngquant.org/) via its knitr hook for compression.
This change enables future scaling of the gallery, as it
eliminates the `R CMD check` note when installed size > 5Mb.
See [#16](https://github.com/nanxstats/ggsci/issues/16)
for more technical details.

Starting with this release, I have switched to a proper, three-component
version number (x.y.z) following the Semantic Versioning recommendations.
Previously, I have been using x.y only without the z component in every
version, which is interpreted by CRAN as equivalent to x.y.0.

See the [ggsci changelog](https://nanx.me/ggsci/news/) for the full list of changes.

## Acknowledgments

A big shout out to [Joshua H. Cook](https://github.com/jhrcook)
and [Clara Jégousse](https://github.com/clarajegousse) for identifying
and contributing the new color scales and
[Dan Chaltiel](https://github.com/DanChaltiel)
for the pro tip on global options.
A special thanks to the users (8 million total downloads)
and package developers (16 reverse imports, 7 reverse suggests)
for your consistent support and trust.
