---
title: "Parsing Human-Readable Text Data from the Web with Readbility.js and R"
author: "Nan Xiao"
date: "2022-08-02T01:30:00"
slug: "r-readability-parser"
categories: []
tags:
  - R
  - JavaScript
  - Readability
  - V8
meta_img: "image/nick-hillier-IEkMMvdZFc0-unsplash.jpg"
description: "In this post, I show how to implement an R wrapper for the JavaScript library Readability.js using the R package V8. The parser allows extracting human-readable text data from any web page."
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  message = FALSE,
  eval = FALSE
)
```

> The R and JavaScript code to reproduce the results in this post is available from
> <https://github.com/nanxstats/r-readability-parser>.

![Photo by [Nick Hillier](https://unsplash.com/photos/IEkMMvdZFc0).](/image/nick-hillier-IEkMMvdZFc0-unsplash.jpg)

## Readability.js

Maybe you have used tools like [rvest](https://rvest.tidyverse.org/) to
harvest **text data** from web pages. Naturally, this often requires
elaborated human efforts in the front to understand the structure of
the target website.

The picture looks quite different when we think at the web scale.
To parse the content of many more sites and many more types of pages,
we need to make our tool adaptive enough to extract the most relevant
text instead of purely relying on manually crafted logic.
We might tolerate missing some useful text and including some
irrelevant text, which is acceptable because they probably won't matter
when the text data we collect is big enough.

Fortunately, [Readability.js](https://github.com/mozilla/readability)
offers a tool for parsing human-readable text from any web page.
It was built for the [Reader View](https://support.mozilla.org/en-US/kb/firefox-reader-view-clutter-free-web-pages) feature in Firefox but
is also usable as an open source, standalone JavaScript library.

In this post, I will create an R wrapper for Readability.js using the
R package [V8](https://github.com/jeroen/V8).

## Packing JS dependencies

Before we write the wrapper, the first step is identifying and packing the
JavaScript dependencies to run in the V8 engine. The three key dependencies
are [@mozilla/readability](https://www.npmjs.com/package/@mozilla/readability),
[jsdom](https://www.npmjs.com/package/jsdom),
and [dompurify](https://www.npmjs.com/package/dompurify).

Following the vignette
[using NPM packages in V8](https://cran.r-project.org/web/packages/V8/vignettes/npm.html),
we pack them as follows.

```{bash}
brew install node
npm install -g browserify
```

Pack Readability.js:

```{bash}
npm install @mozilla/readability
echo "window.Readability = require('@mozilla/readability');" > in.js
browserify in.js -o readability.js
```

Pack jsdom for converting HTML into operable DOM document objects:

```{bash}
npm install jsdom
echo "window.jsdom = require('jsdom');" > in.js
browserify in.js -o jsdom.js
```

Pack DOMPurify mentioned in the readability.js
[security recommendation](https://github.com/mozilla/readability#security)
for sanitizing output to avoid script injection:

```{bash}
npm install dompurify
echo "window.dompurify = require('dompurify');" > in.js
browserify in.js -o dompurify.js
```

## Writing an R binding

We will write some wrapper JavaScript functions to implement the workflow
that uses all three JS libraries above.

```{js}
function readabilityParser(html, url, candidates, threshold) {
  // Parse jsdom with readability.js
  let doc = new jsdom.JSDOM(
    html,
    { url: url }
  );
  let reader = new Readability.Readability(
    doc.window.document,
    { nbTopCandidates: candidates, charThreshold: threshold }
  );
  let res = reader.parse();

  // Sanitize results to avoid script injection
  const purifyWindow = new jsdom.JSDOM('').window;
  const DOMPurify = dompurify(purifyWindow);

  let clean = DOMPurify.sanitize(res.content);
  res.content = clean;

  return res;
}

function isReadable(html, min_content_length, min_score) {
  let doc = new jsdom.JSDOM(html);
  return Readability.isProbablyReaderable(
    doc.window.document,
    { minContentLength: min_content_length, minScore: min_score }
  );
}
```

The R wrapper is quite straightforward if you follow the
[V8 introduction vignette](https://cran.r-project.org/web/packages/V8/vignettes/v8_intro.html).
As is suggested, the interactive JavaScript console via `ct$console()`
is both fun and useful to play with when debugging.

```{r}
readability <- function(html, url, candidates = 5L, threshold = 500L) {
  ct <- V8::v8(global = "window")

  ct$eval("function setTimeout(){}")
  ct$eval("function clearInterval(){}")
  ct$source("js/encoding.min.js")
  ct$source("js/jsdom.js")
  ct$source("js/dompurify.js")
  ct$source("js/readability.js")
  ct$eval(readLines("js/readability-parser.js"))

  # ct$get(V8::JS("Object.keys(window)"))
  ct$call("readabilityParser", html, url, candidates, threshold)
}

is_readable <- function(html, min_content_length = 140, min_score = 20) {
  ct <- V8::v8(global = "window")

  ct$eval("function setTimeout(){}")
  ct$eval("function clearInterval(){}")
  ct$source("js/encoding.min.js")
  ct$source("js/jsdom.js")
  ct$source("js/readability.js")
  ct$eval(readLines("js/readability-parser.js"))

  # ct$get(V8::JS("Object.keys(window)"))
  ct$call("isReadable", html, min_content_length, min_score)
}
```

## Example

Let's parse a recipe page
([pasta with caramelized peppers, anchovies, and ricotta](https://cooking.nytimes.com/recipes/1021246-pasta-with-caramelized-peppers-anchovies-and-ricotta))
from NYT Cooking.

```{r, echo=FALSE, eval=TRUE}
source("readability.R")
```

Check if it is likely that the page is suitable for readability parsing:

```{r, eval=TRUE}
url <- "https://cooking.nytimes.com/recipes/1021246-pasta-with-caramelized-peppers-anchovies-and-ricotta"

html <- url |>
  rvest::read_html() |>
  as.character()

html |> is_readable()
```

We can get the title and the clean, plain text corpus, usable for
downstream text data modeling:

```{r, eval=TRUE}
lst <- html |> readability(url = url)
cat(lst$title)
```

```{r}
lst$textContent |>
  gsub("\\n", " ", x = _, perl = TRUE) |>
  gsub("^\\s+|\\s+$|\\s+(?=\\s)", "", x = _, perl = TRUE) |>
  stringr::str_wrap(width = 74) |>
  cat()
```

<details>
<summary>**Click here to expand the output**</summary>

```{css, echo=FALSE, eval=TRUE}
details > summary {
    border: 1px solid #6c757d;
    border-radius: 0.25rem;
    padding: 1rem;
}
```

```{r, echo=FALSE, eval=TRUE}
lst$textContent |>
  gsub("\\n", " ", x = _, perl = TRUE) |>
  gsub("^\\s+|\\s+$|\\s+(?=\\s)", "", x = _, perl = TRUE) |>
  stringr::str_wrap(width = 74) |>
  cat()
```

</details>

<p></p>

We also got the clean HTML that preserves more structural information.
We can process it further, for example, using xml2 or pandoc.

```{r}
lst$content |>
  htmltools::HTML() |>
  htmltools::browsable()
```

```{r, echo=FALSE, eval=TRUE}
lst$content |>
  htmltools::HTML() |>
  htmltools::save_html(file = "example.html", lang = "en")
```

You can preview the [clean HTML here](example.html).

## Common issues

I encountered and resolved two common issues when using the JS libraries.

### TextEncoder is not defined

I used the hints
[here](https://gist.github.com/tim-salabim/a9740baae9ab879b0c79893132d959a3)
and saved [text-encoding](https://cdn.jsdelivr.net/npm/text-encoding@0.7.0/lib/encoding.min.js)
explicitly as another dependency. Doing this will eliminate the error
`ReferenceError: TextEncoder is not defined`
when sourcing `jsdom.js` with `ct$source()`.

### setTimeout/clearInterval is not defined

It seems some web APIs are not available in the V8 standard library.
I followed the suggestions [here](https://github.com/jeroen/V8/issues/18)
and defined stubs for `setTimeout()` and `clearInterval()`
to avoid errors like `ReferenceError: setTimeout is not defined`
when running jsdom.
