---
title: "Honey, I Did Serialization in 83 R Versions"
author: "Nan Xiao"
date: "2022-05-01T20:00:00"
slug: r-serialize-timemachine
categories: []
tags:
  - R
  - serialization
  - External Data Representation
  - XDR
  - innoextract
meta_img: "image/alex-o9isBQ25H-g-unsplash.jpg"
description: "Look into the R serialization format in all R releases since 1.9.1."
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  message = FALSE,
  echo = FALSE
)
```

> The R code to reproduce the results in this post is available from
> <https://github.com/nanxstats/r-serialize-timemachine>.

![Photo by [Alex Gogan](https://unsplash.com/photos/o9isBQ25H-g).](/image/alex-o9isBQ25H-g-unsplash.jpg)

## A mystery on serialize()

Serialization/deserialization is an important topic for exchanging data
efficiently at scale. In R, there is a native choice for this:
`serialize()`/`unserialize()` and their more convenient interface
`saveRDS()`/`readRDS()`.

Yihui once [asked](https://d.cosx.org/d/419804-serialize-17) why the
first 14 bytes in R serialized data were skipped in `digest::digest()`,
instead of the first 17 bytes for the binary format,
as the additional three filling zero-bytes are always there.

Although there is an entire section in _R Internals_ about
[serialization formats](https://cran.r-project.org/doc/manuals/r-devel/R-ints.html#Serialization-Formats),
I did not find any detailed technical explanations about the bytes in the header.
So I decided to collect more empirical evidence to answer the question.

## An unlikely solution

My first assumption is that seeing the same data serialized in different
R versions instead of different data serialized in the same R version
might give us more information.
This is because the non-data-encoding section in the header likely only changes
when the R versions are different, which will make any minor variations
more observable and thus more interpretable.

This solution then becomes a pure automation exercise.
To maximize the number of R versions I can test,
we need to choose the right platform.

- We should avoid compiling from source because it is almost
  impossible to reuse the original toolchains after
  so many years. Using compiled R binaries would be our best bet.
- To run all the previously compiled R binaries on a single, modern platform,
  we will want to choose Windows because it has probably the best
  [ABI compatibility](https://en.wikipedia.org/wiki/Binary-code_compatibility)
  among the common platforms.

It eventually took ~130 lines of R code to accomplish this automation.
The project is available at
<https://github.com/nanxstats/r-serialize-timemachine>.
You can click the button below to view the serialization results.

<details>
<summary>**Click here to expand the table**</summary>

<!-- <https://yihui.org/en/2021/06/css-full-width/> -->

```{css}
.fullwidth table {
    max-width: none;
    width: 100vw;
    margin-left: calc(50% - 50vw);
    font-family: ui-monospace, SFMono-Regular, "SF Mono", Menlo,
        Consolas, "Liberation Mono", monospace;
    font-size: 0.875rem;
}

details > summary {
    border: 1px solid #6c757d;
    border-radius: 0.25rem;
    padding: 1rem;
}
```

::: {.fullwidth}
```{r}
rds <- stringr::str_sort(fs::dir_ls("output/"), numeric = TRUE) # Natural sorting
version <- gsub(pattern = "R-|.rds", replacement = "", x = fs::path_file(rds))

get_file_size <- function(file) {
  file.info(file)$size
}

readbin_to_char <- function(path) {
  as.character(readBin(path, what = "raw", n = get_file_size(path)))
}

k <- length(version)
hex <- rep(NA, k)
for (i in seq_len(k)) hex[i] <- paste(readbin_to_char(rds[i]), collapse = " ")
df <- data.frame("version" = version, "hex" = hex)

knitr::kable(
  df,
  format = "html",
  table.attr = "class=\"table table-sm table-striped table-hover\"",
  col.names = c("R Version", "Hex value of serialized \"ABCDEF\"")
)
```
:::
</details>

<p></p>

## An evolving language

From this small window, we can have a glimpse at how the infrastructure in R
evolved in the last 20 years, tracing from the earliest release we can test
(R 1.9.1, released in 2004):

- The differences in the serialized data since R 3.6.0 are apparent.
  If you still remember, it was because the
  [serialization format version 3 became the default](https://cran.r-project.org/doc/manuals/r-release/NEWS.3.html#:~:text=CHANGES%20IN%20R%203.6.0),
  although it has already existed since R 3.5.0.
- There are notable differences in R 4.2.0, although still using serialization
  format version 3. Perhaps this is related to the
  [UCRT update](https://developer.r-project.org/Blog/public/2021/12/07/upcoming-changes-in-r-4.2-on-windows/)?
- `serialize()` return value. We cannot use `serialize(connection = NULL)`
  as our test payload directly since [it returned a character string](https://cran-archive.r-project.org/bin/windows/base/old/2.4.0/NEWS.R-2.4.0)
  instead of a raw vector until R 2.4.0.
  Therefore, we used the higher-level function `saveRDS()`
  as a proxy to get the outputs.
- `saveRDS()` compression option. For our purpose of cross-version comparison,
  we set `saveRDS(compress = FALSE)` because the default of `compress` was
  flipped to `TRUE` since R 2.4.0.
- `saveRDS()` was called `.saveRDS()` before R 2.13.0.
- `Rscript.exe` did not exist until R 2.5.0. Therefore, we used `Rcmd.exe`
  instead in the earlier versions.

I think these are all very positive language and tooling improvements---which
benefit all R developers every day!
The consistency and compatibility in other aspects are also amazingly high.
If we don't remove each R version after they are extracted into `dist/`,
you can open them and run every `app/bin/Rgui.exe` on the latest
Windows 10 without issues.

## A possible answer

Here is my answer to the original question on why the skipping offset
should be 14 instead of 17.

From the table above, there are many `00` as zero-bytes of fills.
So naturally, it is critical to know how these filler bytes are used.
If we look into the `serialize()` upstream serialization format
[XDR](https://en.wikipedia.org/wiki/External_Data_Representation),
its corresponding RFC 1832 offers an informative
[example](https://datatracker.ietf.org/doc/html/rfc1832#section-6)
and some
[clues](https://datatracker.ietf.org/doc/html/rfc1832#section-2):

> **BASIC BLOCK SIZE**
>
> The representation of all items requires a multiple of four bytes (or 32 bits) of data.
> ... If the n bytes needed to contain the data are not
> a multiple of four, then the n bytes are followed by enough (0 to 3)
> residual zero bytes, r, to make the total byte count a multiple of 4.

Using R 4.2.0 as an example, the serialized `"ABCDEF"` is:

```text
58 0a
00 00 00 03
00 04 02 00
00 03 05 00
00 00 00 05
...
```

We can annotate it like this:

```text
OFFSET      HEX BYTES       ASCII    COMMENTS
------      ---------       -----    --------
     0      58 0a           X\n      -- X (XDR format) and line break
     2      00 00 00 03     ...3     -- serialization format version = 3
     6      00 04 02 00     .420     -- current R version = 4.2.0
    10      00 03 05 00     .350     -- format available since 3.5.0
    14      00 00 00 05     ...5     -- serialized data starting
```

This is a rough hypothesis, and I could be wrong.
So, don't be shy and leave a comment to add the correct explanation.
