---
title: "Żmij for R: Floating-point to string conversion with round-trip guarantees"
author: "Nan Xiao"
date: "2026-08-09T22:21:00"
slug: zmij-r
categories: []
tags:
  - R
  - R packages
  - Rust
  - extendr
  - plaintext workflow
  - language interoperability
meta_img: "image/claire-satera-gd21U8VUsLc-unsplash.png"
description: >
  zmij converts doubles to the shortest decimal strings that parse back
  to exactly the same values. It brings Victor Zverovich's fast dtoa
  algorithm to R via Rust and extendr.
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  message = FALSE,
  collapse = TRUE
)
options(width = 40)
```

![Photo by [Claire Satera](https://unsplash.com/photos/gd21U8VUsLc).
Stylized with Paper Shaders halftone dots filter.
](/image/claire-satera-gd21U8VUsLc-unsplash.png)

<!--
https://shaders.paper.design/halftone-dots

import { HalftoneDots } from '@paper-design/shaders-react';

<HalftoneDots
  width={1280}
  height={720}
  image="https://paper.design/flowers.webp"
  colorBack="#f2f1e8"
  colorFront="#0030a8"
  originalColors={false}
  type="gooey"
  grid="hex"
  inverted={false}
  size={0.15}
  radius={0.99}
  contrast={0.42}
  grainMixer={0.2}
  grainOverlay={0.2}
  grainSize={0.5}
  fit="contain"
/>
-->

I'm glad to share that [zmij](https://github.com/nanxstats/zmij)
is now on CRAN. zmij converts doubles to decimal strings and back,
using the fewest digits needed to recover the original values exactly.
It brings Victor Zverovich's [zmij algorithm](https://github.com/vitaut/zmij)
to R, via David Tolnay's [Rust zmij port](https://github.com/dtolnay/zmij),
wrapped in a vectorized interface with [extendr](https://extendr.rs/).

You can install it from CRAN with:

```r
install.packages("zmij")
```

## Why bother?

I like storing data as plain text when possible: humans can read it, diff it,
put it under version control, and edit it with any text editor.
Floating-point numbers make this harder than it should be.
By default, R converts them to strings using 15 significant digits
and quietly loses information:

```{r}
x <- c(pi, 0.1, 1 / 3)
x |> as.character()
x |> as.character() |> as.numeric() |> identical(x)
```

You can round-trip reliably by asking for 17 significant digits, but
then you get to look at this:

```{r}
x |> sprintf(fmt = "%.17g")
x |> sprintf(fmt = "%.17g") |> as.numeric() |> identical(x)
```

So the traditional choice was between losing precision, bloating your
files with digit noise, or giving up on plaintext workflow entirely and
using a binary format like `.rds`, `.parquet`, or `.safetensors`.

The dtoa (double to ASCII) algorithms eliminate this trade-off.
For every floating-point number, it finds the shortest decimal string
that parses back to exactly the same value: `0.1` stays `"0.1"`, while `pi`
keeps all 16 digits it genuinely needs. At the moment, zmij also happens to be
the fastest known algorithm to do so with proven correctness:
Victor's [blog post](https://vitaut.net/posts/2025/faster-dtoa/)
has the benchmarks and the tricks behind them. From R, crossing the
language boundary will absorb most of those nanoseconds, which I can
live with: the round-trip guarantee is what I came for.

## Usage

zmij provides two functions. `format_double()` turns doubles into
strings, and `parse_double()` reverses it:

```{r}
library(zmij)
```

```{r}
x <- c(pi, 0.1, 1 / 3, .Machine$double.xmax)
x |> format_double()
x |> format_double() |> parse_double() |> identical(x)
```

That `TRUE` holds for every finite double, meaning your numbers can
survive a trip through any text file untouched:

```{r}
path <- tempfile()
x |> format_double() |> writeLines(path)
path |> readLines() |> parse_double() |> identical(x)
```

Both functions are vectorized and preserve names and dimensions, so
matrices and arrays pass through with their shape intact. Missing and
non-finite values get explicit representations:

```{r}
c(NA_real_, NaN, Inf, -Inf) |> format_double()
```

`parse_double()` is useful in its own right, beyond undoing
`format_double()`: it uses Rust's correctly rounded parser, which
behaves consistently across platforms even at the far edges of the
double range:

```{r}
c("5e-324", "1.7976931348623157e+308") |> parse_double()
```

## Development notes

zmij is my first package built with extendr. It scaffolds a Rust-powered
R package and handles the plumbing between the two languages.
The experience was pleasantly boring: most of my time went into building the
Rust wrapper logic, and almost none into the bridge or packaging.

One piece of developer tooling deserves a mention.
My [r-rust-pkgs](/blog/post/r-rust-pkgs/) project tracks every CRAN package
that uses Rust, and it now comes with a script that turns that list into
something directly useful:

```bash
git clone https://github.com/nanxstats/r-rust-pkgs.git
cd r-rust-pkgs
Rscript scripts/monorepo.R
```

This downloads the source of all those packages into a single
"synthetic monorepo" under `references/`. Move that directory into your
own package (remembering to add it to `.gitignore` and `.Rbuildignore`)
and you have a greppable record of how the entire Rust-in-R community
has solved the problems you might hit. AI coding agents can
search it too. When zmij failed to build on win-builder, a quick
grep showed other packages had already patched their `Makevars`
templates for the same issue, and the fix
([nanxstats/zmij#6](https://github.com/nanxstats/zmij/pull/6))
only took a few minutes to implement.

Thanks to Victor Zverovich for the algorithm and to David Tolnay for
the Rust crate. If you find a use for zmij, or a number it mishandles,
please [create an issue](https://github.com/nanxstats/zmij/issues).
