---
title: "More Than 1,500 File Extensions Are Used Inside R Packages"
author: "Nan Xiao"
date: "2021-12-01"
slug: cran-file-exts
categories: []
tags:
  - R
meta_img: image/tomatoes-pattern.jpg
description: "A glimpse of the file extensions used in R packages."
---

> The R code to reproduce the results is available from the GitHub repo
> [nanxstats/cran-file-exts](https://github.com/nanxstats/cran-file-exts).

```{r, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, echo = TRUE, message = FALSE)
options(width = 70)
```

![Photo by [Pawel Czerwinski](https://unsplash.com/photos/3T4rfR_XZf0).](/image/tomatoes-pattern.jpg)

When applied correctly, file extensions can be informative.
They are the very first clue on handling a specific file
without parsing the file content.

To properly [capture](https://github.com/Merck/pkglite/issues/20) and
[classify](https://github.com/Merck/pkglite/issues/18) files in
source R packages, I am interested in learning what file extensions
are frequently used by R packages.

We can achieve this easily by downloading all R packages available
from CRAN one at a time and collect the file extensions inside:

```{r, eval=FALSE}
library("curl")
library("tools")

repo <- "https://cran.rstudio.com/"
db <- as.data.frame(available.packages(paste0(repo, "src/contrib/")), stringsAsFactors = FALSE)
pkgs <- db$Package
files <- paste0(pkgs, "_", db$Version, ".tar.gz")
links <- paste0(repo, "src/contrib/", files)

find_ext <- function(path) {
  x <- unique(file_ext(untar(path, list = TRUE)))
  x[!(x %in% "")]
}

for (i in seq_along(pkgs)) {
  cat("Downloading", i, "/", length(pkgs), "\n")
  curl_download(links[i], destfile = files[i])
  x <- find_ext(files[i])
  write(paste0(x, collapse = "\t"), file = "exts.txt", append = TRUE)
  unlink(files[i])
}
```

Since this is very one-dimensional, we should look into the frequency table:

```{r}
x <- readLines("exts.txt")
x <- tolower(unlist(strsplit(x, split = "\t")))
y <- sort(table(x), decreasing = TRUE)
```

It looks like we have 1,529 file extensions.
It is also likely a heavy-tailed distribution,
with 96\% of all files designated 5\% of the unique file types.

```{r}
length(y)
z <- y[y > 50L]
length(z) / length(y)
sum(z) / sum(y)
```

We can also cluster this frequency data with any one-dimensional data
clustering algorithm such as the maximum homogeneity clustering,
implemented in my R package [oneclust](https://nanx.me/oneclust/).
Say, we are interested in file extensions that appeared >=5 times:

```{r}
library("oneclust")

eoi <- y[y > 4L]
cl <- oneclust(eoi, 4)
cl$cluster
```

Create a table for everything and display it with the awesome
[DT](https://rstudio.github.io/DT/):

```{r}
df <- data.frame(
  "ext" = names(eoi),
  "mime" = mime::guess_type(paste0(".", names(eoi))),
  "count" = as.vector(eoi),
  "cluster" = dplyr::recode(cl$cluster, `1` = 4, `2` = 3, `3` = 2, `4` = 1)
)
```

After looking into the table, what is **your** interesting discovery?

<style>
div.dataTables_info, div.dataTables_paginate { font-size: 12px; }
table { font-variant-numeric: tabular-nums; }
</style>

```{r, echo=FALSE}
library("DT")

DT::datatable(
  df,
  rownames = FALSE,
  colnames = c("Extension", "Media Type", "Count", "Cluster")
) |>
  formatStyle(
    "cluster",
    color = styleEqual(
      unique(df$cluster), oneclust::cud()[unique(df$cluster)]
    )
  ) |>
  formatStyle(columns = c(1, 2, 3, 4), fontSize = "16px") |>
  formatStyle("mime", fontSize = "12px")
```
