---
title: "Fixing the 'non-numeric argument to binary operator' Error in R Markdown and Quarto"
author: "Nan Xiao"
date: "2023-06-06T20:30:00"
slug: knitr-chunk-option-error
categories: []
tags:
  - R
  - R Markdown
  - Quarto
  - knitr
meta_img: "image/steve-tsang-OZjypvcu3uo-unsplash.jpg"
description: >
  This post explains a common but confusing error message from knitr when
  rendering R Markdown or Quarto documents where a character value is
  assigned to a chunk option that only accepts numeric values.
---

![Rubber duck debugging. Photo by [Timothy Dykes](https://unsplash.com/photos/LhqLdDPcSV8).](/image/timothy-dykes-LhqLdDPcSV8-unsplash.jpg)

If Google leads you here, you probably assigned a character value
to a knitr chunk option such as `fig.width` or `fig.height`
that only accepts numeric values, and your are trying to render
that R Markdown or Quarto document.

The exact error message you encountered looks like this:

```text
Error in `options[[sprintf("fig.%s", i)]] * options$dpi`:
! non-numeric argument to binary operator
Backtrace:
  1. rmarkdown::render("example.Rmd", encoding = "UTF-8")
  2. knitr::knit(knit_input, knit_output, envir = envir, quiet = quiet)
  3. knitr:::process_file(text, output)
  8. knitr:::process_group.block(group)
  9. knitr:::call_block(x)
 10. knitr:::fix_options(params)
 11. options[[o <- sprintf("out.%s", i)]] %n% ...
```

This is because it is easy to make this mistake:

````{verbatim, lang="r"}
```{r, fig.width="100%"}
plot(1:10)
```
````

where you probably meant:

````{verbatim, lang="r"}
```{r, fig.width=6, out.width="100%"}
plot(1:10)
```
````

That's all! I hope this helps.
