---
title: "Sparse Index Tracking with msaenet and CVXR: A Two-Stage Regression Approach"
author: "Nan Xiao"
date: "2023-05-31T01:00:00"
slug: sparse-index-tracking
categories: []
tags:
  - sparse index tracking
  - sparse portfolio
  - sparse linear models
  - regression
  - variable selection
  - convex optimization
  - R
meta_img: "image/kenny-eliason-SvhXD3kPSTY-unsplash.jpg"
description: >
  Explore the two-stage regression approach for sparse index tracking
  with the R packages msaenet and CVXR. Uncover insights into creating
  simpler and more parsimonious models.
---

> The code to reproduce the results in this post is also available
> [here](https://gist.github.com/nanxstats/43ef32acbff6e946637fdda24a052710).

```{r, include=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  echo = TRUE,
  message = FALSE,
  out.width = "95%",
  cache = TRUE
)
```

![The illusion of choice. Photo by [Kenny Eliason](https://unsplash.com/photos/SvhXD3kPSTY).](/image/kenny-eliason-SvhXD3kPSTY-unsplash.jpg)

```{css, echo=FALSE}
.callout {
    margin-top: 1.25rem;
    margin-bottom: 1.25rem;
    padding-top: 1rem;
    padding-left: 1rem;
    border-radius: 0.25rem;
    border-left: 5px solid #64748b;
    border-right: 1px solid #dee2e6;
    border-top: 1px solid #dee2e6;
    border-bottom: 1px solid #dee2e6;
}
```

::::: {.callout}
## Disclaimer

The content in this blog post is for educational and research purposes only.
It is not intended to be investment advice, and the author is not a licensed
financial advisor. Any investment decisions should be based on your own
analysis and consultation with a licensed financial advisor.
The author is not responsible for any financial losses or damages
resulting from the use of this information.
:::::

## Background

I always appreciate recommender systems algorithms as they solve the
information overload problem, which is quite prevalent in modern life.
Sparse index tracking aims to use a limited number of equities to track a
financial index. This approach addresses a similar problem as portfolio
selection and optimization, with further implications for balancing
investment returns and controlling risk.

This topic of sparse portfolios has been discussed in, for example,
[Brodie et al. (2009)](https://doi.org/10.1073/pnas.0904287106) and
[Benidis, Feng, and Palomar (2018)](https://doi.org/10.1109/TSP.2017.2762286).
In the Benidis paper and their amazing R package
[sparseIndexTracking](https://cran.r-project.org/package=sparseIndexTracking),
the sparse index tracking problem was formulated as a constrained,
$\ell_0$-norm regression:

$$
\begin{array}{c}
\min_\limits{\beta} || y - X\beta ||_2^2 + \lambda ||\beta||_0\\
\text{s.t.} \sum \beta_i = 1, \beta_i >= 0\\
\end{array}
$$

where $y$ can be the daily returns of the index we are tracking,
$X$ is the daily returns of the (large number of) assets,
and $\beta$ being the parameters to be estimated.
We can also optimize for different types of tracking errors in this framework.

In the experiments demonstrated by the package on a dataset of S&P 500
returns of year 2010, they were able to use 40--50 assets to track the index.
I am always curious about two things, though.
First, can we have a even sparser portfolio? Since 40 assets
still feels a lot to me.
Second, whether we could use a simpler two-step approach (optimization-wise)
briefly described in their paper, where one could do asset
selection and capital allocation separately.
At the end of the analysis below, my conclusion is that both of these
are possible, but not without potential trade-offs,
at least in terms of tracking error.

## Data

The dataset is S&P 500 returns in year 2010 used in the sparseIndexTracking
[vignette](https://cran.r-project.org/web/packages/sparseIndexTracking/vignettes/SparseIndexTracking.html).
Load the data and split into a training and a test set:

```{r}
library("xts")

index2010 <- sparseIndexTracking::INDEX_2010
x_tr <- index2010$X[1:126]
x_te <- index2010$X[127:252]
r_tr <- index2010$SP500[1:126]
r_te <- index2010$SP500[127:252]
```

Create wrapper functions for P&L plotting using ggplot2, cowplot,
and [ggsci](https://nanx.me/ggsci/):

```{r}
df_pnl <- function(x_te, beta, y_test, title) {
  df <- cbind(
    cumprod(1 + x_te %*% beta),
    cumprod(1 + y_test)
  )
  names(df)[1] <- title
  df
}

plot_pnl <- function(object, title = "Cumulative P&L") {
  ggplot2::autoplot(object, facets = NULL) +
    ggplot2::ggtitle(title) +
    cowplot::theme_minimal_hgrid() +
    ggsci::scale_color_d3() +
    ggplot2::scale_y_continuous(breaks = seq(1, 1.5, .05), limits = c(1, 1.4)) +
    ggplot2::scale_x_date(date_breaks = "1 months", date_labels = "%b") +
    ggplot2::theme(
      axis.title.x = ggplot2::element_blank(),
      legend.position = "bottom"
    )
}
```

## Previous models

The original vignette used $\ell_0$ penalty and optimized for four
different tracking measures: empirical tracking error,
downside risk, Huber empirical tracking error, and Huber downside risk.
I will only show the empirical tracking error model here:

```{r}
w_ete <- sparseIndexTracking::spIndexTrack(
  x_tr,
  r = r_tr, lambda = 1e-7, u = 0.5, measure = "ete"
)

names(w_ete[w_ete > 1e-6])
```

<details>
<summary>Click here to expand the detailed portfolio</summary>

```{css, echo=FALSE, eval=TRUE}
details > summary {
    border: 1px solid #cbd5e1;
    border-radius: 0.25rem;
    padding: 1rem;
}
```

```{r}
data.frame(sort(w_ete, decreasing = TRUE), fix.empty.names = FALSE)
```

</details>

<p></p>

```{r, fig.cap="ETE portfolio and SP500 cumulative P&L."}
df_pnl(x_te, w_ete, r_te, "PortfolioETE") |> plot_pnl()
```

## A two-step algorithm for parsimonious solutions

Revisiting my questions: can we get an more parsimonious model with
fewer assets selected while still maintaining tracking performance relatively well?
Additionally, instead of optimizing a constrained sparse regression objective,
we could potentially use a two-stage algorithm.
Following these ideas, things should work like this:

1. Use an unconstrained sparse regression for selecting the assets,
   preferably yielding sparser solutions.
2. Use a constrained OLS regression on the selected assets where the
   coefficients are non-negative and sum to 1.

This is obviously a "poor man's approach" but is simpler in terms of
implementation as even I can do it in no time, and it's shown below.

### Stage 1

Create wrapper functions for extracting $\beta$s from
[msaenet](https://nanx.me/msaenet/) models:

```{r}
get_beta <- function(object) {
  beta <- coef(object)
  names(beta) <- if (inherits(object, "msaenet.msaenet")) {
    rownames(object$beta)
  } else {
    colnames(object$model$X)
  }
  beta
}

get_nzv <- function(object) {
  beta <- get_beta(object)
  names(beta[which(beta != 0)])
}
```

### Stage 2

Use CVXR by [Fu et al. (2020)](https://doi.org/10.18637/jss.v094.i14)
to do a constrained sum-to-one OLS regression on
the selected variables to get the weights:

```{r}
library("CVXR")

storeg <- function(x, y) {
  x <- as.matrix(x)
  y <- as.vector(y)
  p <- ncol(x)
  beta <- Variable(p)

  obj <- sum((y - x %*% beta)^2)
  constr <- list(beta >= 0, sum(beta) == 1)
  prob <- Problem(Minimize(obj), constr)
  result <- solve(prob)

  structure(
    list(
      result = result,
      beta = result$getValue(beta)
    ),
    class = "storeg"
  )
}
```

Get asset names and coefficients in the portfolio:

```{r}
get_portfolio <- function(beta, nzv) {
  as.vector(beta) |>
    setNames(nzv) |>
    sort(decreasing = TRUE) |>
    data.frame(fix.empty.names = FALSE)
}
```

Chaining things together:

```{r, fig.cap="msaenet portfolio and SP500 cumulative P&L."}
fit_enet <- msaenet::msaenet(
  x = as.matrix(x_tr),
  y = as.vector(r_tr),
  family = "gaussian",
  init = "ridge",
  tune = "cv",
  nsteps = 10,
  lower.limits = 0,
  verbose = FALSE,
  seed = 42
)

fit_enet |> get_nzv()

fit_enet_sto <- storeg(x_tr[, get_nzv(fit_enet)], r_tr)

get_portfolio(fit_enet_sto$beta, get_nzv(fit_enet))

df_pnl(x_te[, get_nzv(fit_enet)], fit_enet_sto$beta, r_te, "Portfolio.msaenet") |>
  plot_pnl()
```

We have selected much fewer (five) assets to track the index.
The tracking error was also not too big.

## Stability of the solution

Since we used cross validation to tune the hyperparameters,
there is a possibility that we were "lucky" in getting a good model
so it might be useful to check the stability of the selection results
in many repeats with different seeds, and consequently, varying splits of the
training data.

```{r, eval=FALSE}
library("doParallel")
registerDoParallel(detectCores())

fit_rep <- foreach::foreach(seed = 1:100) %dopar% {
  msaenet::msaenet(
    x = as.matrix(x_tr),
    y = as.vector(r_tr),
    family = "gaussian",
    init = "ridge",
    tune = "cv",
    nsteps = 10,
    lower.limits = 0,
    verbose = FALSE,
    seed = seed
  )
}
```

```{r, eval=FALSE, echo=FALSE}
saveRDS(fit_rep, file = "fit_rep.rds")
```

```{r, echo=FALSE}
fit_rep <- readRDS("fit_rep.rds")
```

```{r, fig.asp=0.25, fig.cap="Dot plot of asset selection frequency out of 100 experiments."}
as.data.frame(table(unlist(sapply(fit_rep, get_nzv)))) |>
  ggplot2::ggplot(ggplot2::aes(x = Freq, y = Var1)) +
  ggplot2::geom_point(size = 3, color = ggsci::pal_d3()(1)) +
  ggplot2::scale_x_continuous(
    name = "Selection frequency out of 100 experiments",
    limits = c(0, 100),
    expand = c(0, 5)
  ) +
  ggplot2::scale_y_discrete(name = NULL, expand = c(0, 0.5)) +
  cowplot::theme_minimal_grid()
```

## Even sparser solutions

We can plug in different penalized regression methods in the first stage to
get different portfolios with different degrees of sparsity and asset structure.
For example, we can try to get an even sparser solution:

```{r, fig.cap="Sparse portfolio using multi-step SCAD-net and SP500 cumulative P&L."}
fit_snet <- suppressWarnings(msaenet::msasnet(
  x = as.matrix(x_tr),
  y = as.vector(r_tr),
  family = "gaussian",
  init = "ridge",
  tune = "cv",
  nsteps = 10,
  verbose = FALSE,
  seed = 42
))

fit_snet |> get_nzv()

fit_snet_sto <- storeg(x_tr[, get_nzv(fit_snet)], r_tr)

get_portfolio(fit_snet_sto$beta, get_nzv(fit_snet))

df_pnl(x_te[, get_nzv(fit_snet)], fit_snet_sto$beta, r_te, "Portfolio.msasnet") |>
  plot_pnl()
```

As we further reduce the number of assets from five to three,
the tracking error was ok in the beginning of the testing period but
went up in the latter part. Similar situation for another try:

```{r, fig.cap="Sparse portfolio using multi-step MCP-net and SP500 cumulative P&L."}
fit_mnet <- suppressWarnings(msaenet::msamnet(
  x = as.matrix(x_tr),
  y = as.vector(r_tr),
  family = "gaussian",
  init = "ridge",
  tune = "cv",
  nsteps = 10,
  verbose = FALSE,
  seed = 42
))

fit_mnet |> get_nzv()

fit_mnet_sto <- storeg(x_tr[, get_nzv(fit_mnet)], r_tr)

get_portfolio(fit_mnet_sto$beta, get_nzv(fit_mnet))

df_pnl(x_te[, get_nzv(fit_mnet)], fit_mnet_sto$beta, r_te, "Portfolio.msamnet") |>
  plot_pnl()
```

## Summary

While there are still open challenges in sparse index tracking such as
determining rebalancing frequency, refining backtest strategies,
and ensuring portfolio stability, this two-stage regression approach
provides a simpler and potentially more parsimonious model for tackling
this problem.
