---
title: "Five Minute Shiny: File Input Area Widget"
author: "Nan Xiao"
date: "2022-04-17T21:00:00"
slug: shiny-file-input-area
categories: []
tags:
  - R
  - Shiny
  - "Bootstrap 5"
meta_img: "image/willian-justen-de-vasconcellos-wnmtuxbOuAs-unsplash.jpg"
description: "Implement a file input area UI widget for Shiny in five minutes."
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  message = FALSE,
  eval = FALSE
)
```

> The R code in this post is also available as [this GitHub Gist](https://gist.github.com/nanxstats/095c4b13eae339faf1783c96390d71cf).

![Photo by [Willian Justen de Vasconcellos](https://unsplash.com/photos/wnmtuxbOuAs)](/image/willian-justen-de-vasconcellos-wnmtuxbOuAs-unsplash.jpg)

Our goal today is to implement a file input **area** widget for Shiny.
Compared to a regular file input button, a larger file input UI helps users
focus on the task where an upload action is the central flow of the web application.

Fortunately, the native
[`shiny::fileInput()`](https://github.com/rstudio/shiny/blob/66968904bfedba03648053094c35b06059f82e63/R/input-file.R)
offers a solid foundation for customization, as it already supports both
click-to-browse and drag-and-drop for file selection.
We only need to move the progress bar, make the button larger,
and apply some custom styles to it.

```{r}
fileInputArea <- function(inputId, label, multiple = FALSE, accept = NULL,
                          width = NULL, buttonLabel = "Browse...", placeholder = "No file selected") {
  restoredValue <- restoreInput(id = inputId, default = NULL)

  # Catch potential edge case - ensure that it's either NULL or a data frame.
  if (!is.null(restoredValue) && !is.data.frame(restoredValue)) {
    warning("Restored value for ", inputId, " has incorrect format.")
    restoredValue <- NULL
  }

  if (!is.null(restoredValue)) {
    restoredValue <- toJSON(restoredValue, strict_atomic = FALSE)
  }

  inputTag <- tags$input(
    id = inputId,
    name = inputId,
    type = "file",
    # Don't use "display: none;" style, which causes keyboard accessibility issue; instead use the following workaround: https://css-tricks.com/places-its-tempting-to-use-display-none-but-dont/
    style = "position: absolute !important; top: -99999px !important; left: -99999px !important;",
    `data-restore` = restoredValue
  )

  if (multiple) {
    inputTag$attribs$multiple <- "multiple"
  }
  if (length(accept) > 0) {
    inputTag$attribs$accept <- paste(accept, collapse = ",")
  }

  div(
    class = "form-group shiny-input-container w-100",
    style = htmltools::css(width = htmltools::validateCssUnit(width)),
    shiny:::shinyInputLabel(inputId, ""),
    div(
      class = "input-group mb-3",
      # input-group-prepend is for bootstrap 4 compat
      tags$label(
        class = "input-group-btn input-group-prepend w-100",
        span(
          class = "btn btn-area w-100", inputTag,
          div(tags$image(src = icon_encoded, width = "80px;"), style = "margin-top: 2rem;"),
          div(p(label), style = "font-size: 1.2rem; font-weight: 700; padding-top: 2rem;"),
          div(p(buttonLabel), style = "font-size: 1rem; font-weight: 400; margin-bottom: 2rem;")
        )
      )
    ),
    tags$div(
      id = paste(inputId, "_progress", sep = ""),
      class = "progress active shiny-file-input-progress",
      tags$div(class = "progress-bar")
    )
  )
}
```

Define the styles for the button and progress bar. Also, include the SVG icon.

```{r}
# Use Bootstrap 5 colors $gray-700 and $gray-600
css_btn_area <- textConnection("
.btn-area {
  color: #495057;
  border-color: #495057;
  border-style: dashed;
  border-width: 2px;
  border-radius: 20px;
  background-color: transparent;
}

.btn-area:hover {
  color: #6c757d;
}

.progress {
  height: 32px;
}

.progress .progress-bar {
  font-size: 16px;
  line-height: 28px;
}")

# Icon from <https://icons.getbootstrap.com/icons/upload/>
icon_file <- tempfile(fileext = ".svg")
writeLines('
<svg xmlns="http://www.w3.org/2000/svg" width="16" height="16" fill="#495057" class="bi bi-upload" viewBox="0 0 16 16">
  <path d="M.5 9.9a.5.5 0 0 1 .5.5v2.5a1 1 0 0 0 1 1h12a1 1 0 0 0 1-1v-2.5a.5.5 0 0 1 1 0v2.5a2 2 0 0 1-2 2H2a2 2 0 0 1-2-2v-2.5a.5.5 0 0 1 .5-.5z"/>
  <path d="M7.646 1.146a.5.5 0 0 1 .708 0l3 3a.5.5 0 0 1-.708.708L8.5 2.707V11.5a.5.5 0 0 1-1 0V2.707L5.354 4.854a.5.5 0 1 1-.708-.708l3-3z"/>
</svg>',
  con = icon_file
)
icon_encoded <- xfun::base64_uri(icon_file)
```

Try it out at <https://nanx.shinyapps.io/shiny-file-input-area/>.

```{r}
card <- function(title, ...) {
  htmltools::tags$div(
    class = "card",
    htmltools::tags$div(class = "card-header", title),
    htmltools::tags$div(class = "card-body", ...)
  )
}

ui <- shiny::fluidPage(
  title = "File input area for Shiny (Bootstrap 5)",
  theme = bslib::bs_theme(version = 5),
  lang = "en",
  includeCSS(css_btn_area),
  shiny::fluidRow(
    style = "margin-top: 20px;",
    shiny::column(
      width = 10, offset = 1,
      card(
        title = "File input area for Shiny (Bootstrap 5)",
        shiny::fluidRow(
          column(
            width = 4, offset = 4,
            fileInputArea(
              "upload",
              label = "Drop your plain text files here!",
              buttonLabel = "Any plain text file, max 100 KB each.",
              multiple = TRUE,
              accept = "text/plain"
            ),
            shiny::tableOutput("files")
          )
        )
      )
    )
  )
)

server <- function(input, output, session) {
  output$files <- renderTable(
    input$upload[, c("name", "type", "size")],
    width = "100%"
  )
}

options(shiny.maxRequestSize = 0.1 * 1024^2)
shiny::shinyApp(ui, server)
```
