---
title: "Five Minute Shiny: Implement the Switch Input with Bootstrap 5"
author: "Nan Xiao"
date: "2022-04-14T00:30:00"
slug: shiny-bs5-switches
categories: []
tags:
  - R
  - Shiny
  - "Bootstrap 5"
meta_img: "image/zonduurzaam-deventer-BFNi3TWB2fw-unsplash.jpg"
description: "Implement the Bootstrap 5 switch input for Shiny in five minutes."
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  message = FALSE,
  eval = FALSE
)
```

> The R code in this post is also available as [this GitHub Gist](https://gist.github.com/nanxstats/e43c1f207dc8b727f6c832d3226d1d16).

![Photo by [Zonduurzaam Deventer](https://unsplash.com/photos/BFNi3TWB2fw).](/image/zonduurzaam-deventer-BFNi3TWB2fw-unsplash.jpg)

Bootstrap 4 and 5 have expanded the types of UI components compared to
version 3. On the other hand, the available UI inputs and outputs in Shiny have
been relatively stable while being carefully made compatible across
different Bootstrap versions through bslib.

Since we can already leverage the native UI components in Bootstrap 5 via bslib,
it would be fun to create the R wrappers for the "missing" inputs that
were not available in Bootstrap 3.
One of the most straightforward examples is the switch input.

A toggle switch input is a checkbox input, with only binary values and
[different use cases](https://uxplanet.org/checkbox-vs-toggle-switch-7fc6e83f10b8).
Therefore, we can borrow `shiny::checkboxInput()` and apply the necessary changes based on
[Bootstrap 5 switches](https://getbootstrap.com/docs/5.1/forms/checks-radios/#switches)
without creating custom input bindings in JavaScript.
To update the switch inputs, we can even reuse `shiny::updateCheckboxInput()`.

```{r}
switchInput <- function(inputId, label, value = FALSE, disabled = FALSE, width = NULL) {
  value <- shiny::restoreInput(id = inputId, default = value)
  inputTag <- htmltools::tags$input(id = inputId, type = "checkbox", role = "switch", class = "form-check-input")
  if (!is.null(value) && value) {
    inputTag$attribs$checked <- NA
  }
  if (!is.null(disabled) && disabled) {
    inputTag$attribs$disabled <- NA
  }
  htmltools::tags$div(
    class = "form-group shiny-input-container", style = htmltools::css(width = shiny::validateCssUnit(width)),
    htmltools::tags$div(
      class = "from-check form-switch",
      inputTag,
      htmltools::tags$label(label, class = "form-check-label")
    ),
  )
}

updateSwitchInput <- shiny::updateCheckboxInput
```

Try it out (preview at <https://nanx.shinyapps.io/shiny-bs5-switches/>):

```{r}
card <- function(title, ...) {
  htmltools::tags$div(
    class = "card",
    htmltools::tags$div(class = "card-header", title),
    htmltools::tags$div(class = "card-body", ...)
  )
}

ui <- shiny::fluidPage(
  title = "Bootstrap 5 Switches for Shiny",
  theme = bslib::bs_theme(
    version = 5,
    primary = "#002FA7",
    `form-check-input-checked-bg-color` = "#002FA7"
  ),
  lang = "en",
  shiny::fluidRow(
    style = "margin-top: 20px;",
    shiny::column(
      width = 8, offset = 2,
      card(
        title = "Bootstrap 5 Switches for Shiny",
        switchInput("input1", label = "Default switch checkbox input", width = "100%"),
        verbatimTextOutput("value1"),
        switchInput("input2", label = "Checked switch checkbox input", value = TRUE, width = "100%"),
        verbatimTextOutput("value2"),
        switchInput("input3", label = "Disabled switch checkbox input", disabled = TRUE, width = "100%"),
        verbatimTextOutput("value3"),
        switchInput("input4", label = "Disabled checked switch checkbox input", value = TRUE, disabled = TRUE, width = "100%"),
        verbatimTextOutput("value4"),
        sliderInput("controller", "Controller", min = 0, max = 4, value = 0, step = 1),
        switchInput("input5", "Input switch that can be updated")
      )
    )
  )
)

server <- function(input, output, session) {
  output$value1 <- renderText(input$input1)
  output$value2 <- renderText(input$input2)
  output$value3 <- renderText(input$input3)
  output$value4 <- renderText(input$input4)
  observe({
    # TRUE if input$controller is odd, FALSE if even.
    x_even <- input$controller %% 2 == 1
    updateSwitchInput(session, "input5", value = x_even)
  })
}

shiny::shinyApp(ui, server)
```
