---
title: "New Paper Published: eCTD Submission with Analysis Using R"
author: "Nan Xiao"
date: "2022-09-28T11:30:00"
slug: ectd-pkglite-paper
categories: []
tags:
  - R
  - R packages
  - pkglite
  - clinical trials
  - regulatory submission
  - reproducible research
meta_img: "image/deepmind-ZJKE4XVlKIA-unsplash.jpg"
description: >
  I am thrilled to share that our paper, Electronic Common Technical Document
  Submission with Analysis Using R, is published online today in Clinical Trials.
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  echo = TRUE,
  message = FALSE
)
```

![Photo by [DeepMind](https://unsplash.com/photos/ZJKE4XVlKIA).](/image/deepmind-ZJKE4XVlKIA-unsplash.jpg)

I am thrilled to share that our paper,
[_Electronic Common Technical Document Submission with Analysis Using R_](https://doi.org/10.1177/17407745221123244),
is published online today in _Clinical Trials_.
We have selected the open access (OA) option to ensure that everyone in the
community can check out the full text freely.

## A simple solution to a complex problem

Our article presented a viable path to prepare proprietary R assets for submission,
following the
[electronic common technical document](https://www.fda.gov/drugs/electronic-regulatory-submission-and-review/electronic-common-technical-document-ectd)
(eCTD) specification.
The goal is to make the submission document self-contained, easy to
create, transfer, restore, review, and submit, with confidence in
transparency and reproducibility for both submitters and reviewers.

The workflow is built upon the idea of organizing the submission assets
as R packages. Our tool, [pkglite](https://github.com/merck/pkglite),
can convert the proprietary source R packages with the analysis code into
a compact text format and restore/install later.
We provided real-world examples and templates, so you can try them out today.
We also highlighted the successful pilot 1 open source submission
from the R Consortium
[R submission working group](https://rconsortium.github.io/submissions-wg/),
leveraging the solution and workflow.
More details of the workflow have been documented in our bookdown project
[_R for Clinical Study Reports and Submission_](https://r4csr.org/).

## A few more words on pkglite

The pkglite package has accumulated 15k total downloads from the
RStudio CRAN mirror.
As a simple, generic text file-based dependency distribution solution for R,
we have observed many other creative use cases for pkglite since it was
first published on CRAN in March 2021.

```{r, fig.cap="Download stats for pkglite, averaging ~800 downloads per month (September 2022).", fig.asp=1/4}
dlstats::cran_stats("pkglite") |>
  ggplot2::ggplot(ggplot2::aes(x = end, y = downloads, color = package)) +
  ggplot2::geom_point() +
  ggplot2::geom_line() +
  cowplot::theme_cowplot() +
  ggsci::scale_color_d3() +
  ggplot2::theme(axis.title.x = ggplot2::element_blank())
```

We hope pkglite can continue to deliver
value to people who need a lightweight R package exchange and distribution solution.
We have some good ideas to further develop the toolchain. You can see
them from my R/Pharma conference talk last year on
[reimagining the R package distribution system for reproducible research and submissions](https://nanx.me/talks/reimagine-rpkgs/).

## Acknowledgments

I would like to thank the editors and the anonymous reviewers for offering
constructive feedback that improved the quality of our manuscript.
I also appreciate the outstanding contributions made by my collaborators
[Yujie Zhao](https://www.linkedin.com/in/yujie-zhao-stat-phd/),
[Keaven Anderson](https://keaven.github.io/),
[Yilong Zhang](https://elong0527.github.io/),
and many excellent colleagues for making this happen as a team. 🤗
